---
title: "Making Choices"
author: "Marcos Padilla-Ruiz"
date: "2022-11-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Logical and conditional statements


```{r}
1 == 1
1 > 1
1 < 1
1 >= 1
1 <= 1

"a" == "a"
"a" < "a"
"a" < "b"
"b" < "c"
"apple" < "banana"
"apple" < "airplane"

c(1, 2, 3, 4, 1, 1) == 1

letters
letters < "m"
letters[letters < "m"]
LETTERS

```

load data from uhuru data set

```{r}
plots <- read.csv(file = "../raw_data/plots.csv")
species <- read.csv(file = "../raw_data/species.csv")
surveys <- read.csv(file = "../raw_data/surveys.csv")

colnames(surveys)
```

```{r}
library(dplyr)
combined <- dplyr::inner_join(surveys, species, by = "species_id")

colnames(combined)
```

extract the year column:
```{r}
colnames(combined)
str(combined)
dim(combined)
combined$year
combined[,4]
combined[,"year"]

combined$year == 1977
combined$genus == "Dipodomys"
data.frame(combined$genus, combined$genus == "Dipodomys")
data.frame(combined$species, combined$species == "merriami")

```

some examples of conditional statements (logical statement AND logical statement we use `&`, OR `|`)
```{r}
1 ==1 & 1 < 2
1 == 1 & 1 == 2
1 == 1 | 1 == 2
```

```{r}
combined$year == 1977 & combined$genus == "Dipodomys" & combined$species == "merriami"

age_class = "sapling"
 if (age_class == "sapling"){
y = 10
 }
 y
 
 age_class = "seedling"
 if (age_class == "sapling"){
y = 10
 } else {
   y = 5
 }
 y

 
 age_class = "adult"
 if (age_class == "sapling"){
   y <- 10
 } else if (age_class == "seedling") {
   y <- 5
 } else {
   y <- 0
 }
 y
```

## using if else inside functions

1. convert your conditionals statments from last exercise into a function called get y that takes age_class as argument and returns y

```{r}
get_y = function(age_class) {
  if (age_class == "sapling") {
   y <- 10 
 } else if (age_class == "seedling") {
   y <-  5 
   } else { 
     y <-  0 
   }
  return(y)
}
```

2. run the function using the values "sapling", "seedling", "adult", "mature", "established"
```{r}
get_y("sapling") # 10
get_y("seedling") # 5
get_y("adult") # 0
get_y ("mature") # 0
get_y("established") # 0
```

## Choice operators

1. Create the following variables:
```{r}
 w <- 10.2
 x <- 1.3
 y <- 2.8
 z <- 17.5
 colors <- c("red", "blue", "green")
 masses <- c(45.2, 36.1, 27.8, 81.6, 42.4)
 dna1 <- "attattaggaccaca"
 dna2 <- "attattaggaacaca"
```
 
2. Use them to print whether or not the following statements are TRUE or FALSE.
```{r}
 w > 10 # a) TRUE
 colors == "green" # b) FALSE FALSE TRUE
 x > y # c) FALSE
 masses > 40 # d) TRUE FALSE FALSE TRUE TRUE
 (2*x+0.2) == y # e) FALSE
 dna1 == dna2 # f) FALSE
 dna1 != dna2 # g) TRUE
 w > x | y > z # h) TRUE
x*w < 13.5 & x*w > 13.2 # i) TRUE
masses < 50 & masses > 30 # j) TRUE TRUE FALSE FALSE TRUE
```


```{r}
get_mass_from_length = function(length, a, b) {
  mass <- a * length^b
}
```  


## Mass estimates

1. Create a new function get_mass_from_length_by_name() that:
takes two arguments, the length and the name of the dinosaur group
```{r}
get_mass_from_length_by_name = function(length, name) {
   if (name == "Stegosauria") {
     a <- 10.95 
     b <- 2.64 
   } else if (name == "Theropoda") {
     a <- 0.73
     b <- 3.63
   } else if (name == "Sauropoda") {
     a <- 214.44
     b <- 1.46
   } else { 
   message("No known estimation for ", name)
   return(NA)
   }
 mass <- get_mass_from_length(length = length, a = a, b = b )
 return(mass)
}  
```

2 & 3. Run the function for:
a) A Stegosauria that is 10 meters long.
b) A Theropoda that is 8 meters long.
c) A Sauropoda that is 12 meters long.
d) A Ankylosauria that is 13 meters long.
```{r}
get_mass_from_length_by_name(10, "Stegosauria") # 4779.848
get_mass_from_length_by_name(8, "Theropoda") # 1385.286
get_mass_from_length_by_name(12, "Stegosauria") # 7734.863

# challenge 3:
get_mass_from_length_by_name(13, "Ankylosauria") # No known estimation for Ankylosauria
```

4. Change your function so that it uses two different values of a and b for Stegosauria:
a) When Stegosauria is greater than 8 meters long use the equation above.
b) When it is less than 8 meters long use a = 8.5 and b = 2.8.
c) Run the function for a Stegosauria that is 6 meters long.
```{r} 
get_mass_from_length_by_name = function(length, name) {
   if (name == "Stegosauria" & length > 8) {
     a <- 10.95 
     b <- 2.64 
    } else if (name == "Stegosauria" & length < 8) {
     a <- 8.5
     b <- 2.8
   } else if (name == "Theropoda") {
     a <- 0.73
     b <- 3.63
   } else if (name == "Sauropoda") {
     a <- 214.44
     b <- 1.46
   } else { 
   message("No known estimation for ", name)
   return(NA)
   }
 mass<- get_mass_from_length(length = length, a = a, b = b )
 return(mass)
}  
get_mass_from_length_by_name(6, "Stegosauria") # 1283.047
```





