---
title: "Creating functions"
author: "Marcos Padilla-Ruiz"
date: "2022-11-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction to creating functions

three reasons why creating your own functions is good practice for data science.

- it allows reusing code for other parts of a project or a future project
- less error prone than copy pasting code
- code is more organized for you and others so it is easier to read
- allows deeper understanding of what your code is
- useful to add data to data frames
- allows getting specific values and requirements
- invites you to be intentional on the code that you are writing
- allows to be more in control of the outputs

## Exercise 1
part 1
```{r}
convert_pounds_to_grams <- function(pounds) {
    grams = 453.6 * pounds
    return(grams)
}

```

part 2
```{r}
convert_pounds_to_grams <- function(pounds = 3.75) {
    grams = 453.6 * pounds
    return(grams)
}

convert_pounds_to_grams() # this requires a default value in function
convert_pounds_to_grams(3.75)
convert_pounds_to_grams(pounds = 3.75)
```


## Exercise 2

part 1
```{r}
get_mass_from_length_theropoda <- function(length){
  mass <- 0.73 * length ^ 3.63
  return(mass)
}

get_mass_from_length_theropoda(16)
```

part 2
```{r}
#1 method
get_mass_from_length <- function(length = 26){
  mass <- 214.44 * length ^ 1.46
  return(mass)
}

get_mass_from_length()

#2 method
get_mass_from_length <- function(length, a, b){
  mass <- a * length ^ b
  return(mass)
}

get_mass_from_length(length = 26, a = 214.44, b = 1.46)
```


## Exercise 3

```{r}
get_mass_from_length <- function(length, a = 39.9, b = 2.6){
  mass <- a * length ^b
  return(mass)
}

get_mass_from_length(length = 22, a = 214.44, b = 1.46)
get_mass_from_length(length = 16, a = 39.9, b = 2.6)
```

## Exercise 4
```{r}
convert_kg_to_pounds <- function(kg) {
  pounds <- 2.205 * kg
}

```


```{r}
library(magrittr)
weight_in_pounds <- get_mass_from_length(12, 10.95, 2.64) %>% 
convert_kg_to_pounds()
weight_in_pounds
```













