---
title: "portal-phylogenies.Rmd"
author: "Marcos Padilla-Ruiz"
date: "2022-10-25"
output:
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# add intro to ape

```{r, eval=FALSE}
install.packages("ape")
```


to read a newick tree from a local file

```{r}
library(ape)
portal_tree <- read.tree(file = "../raw_data/portal-tree.tre")
portal_tree
```

read newick from url

```{r}
small_tree <- read.tree(file = "http://ape-package.ird.fr/APER/APER2/primfive.tre")
small_tree
```


# the structure of a tree in R

```{r}
class(portal_tree)
length(portal_tree)
colnames(portal_tree)
names(portal_tree)
portal_tree$Nnode
portal_tree["tip.label"]
head(portal_tree[[3]])
summary(portal_tree)
str(portal_tree)
class(portal_tree$edge)
portal_tree$edge
```


## plotting phylogenetic trees

```{r}
plot.phylo(x = portal_tree)
```

an extension of phylogenetic visualization: ggtree
to install packages from bioconductor we need a CRAN package called biocmanager

```{r, eval=FALSE}
install.packages("BiocManager")
install("ggtree")
```


```{r}
library(BiocManager)
library(ggtree)
ggtree(portal_tree)
```
# this is equivalent to:

```{r}
ggplot(portal_tree, aes(x, y)) +
  geom_tree() +
  theme_tree()
```


add a scale with the function geom_treescale()
```{r}
ggtree(portal_tree) +
  geom_treescale()
```


exercise: plot the small tree of five species of primates and include a scale
what is the difference in structure between the two trees?

```{r}
ggtree(small_tree) +
  geom_treescale()
```
- A difference in number of tips (43 vs 5)
- our small_tree does not have node.label as opposed to the portal trees but it does include branch lengths through the edge.length element as shown by the tree above


```{r}
names(small_tree)
summary(small_tree)
str(small_tree)
class(small_tree$edge.length)
small_tree$edge.length
```


# add tip labels and node labels

Because a plot is nothing without labels
```{r}
ggtree(portal_tree) +
  geom_treescale() +
  geom_tiplab(size = 1, col = "cyan", fontface = "italic")
```

add a limit to the plot so we can see the labels

```{r}
ggtree(portal_tree) +
  geom_treescale() +
  geom_tiplab(size = 3, col = "springgreen", fontface = "italic") +
  xlim(NA, 90)
```

```{r}
#age of each node
branching.times(small_tree)
```


# homework: make a plot for small_tree with scale, tip labels, and an approprate xlim size to display names fully


```{r}
library(ggtree)
ggtree(portal_tree, layout = "fan", open.angle = 45) -> plot1
```

```{r}
#install.packages("aplot")

library(aplot)
```



```{r}
plot_list(plot1, plot1, labels = c("Circular Tree of Portal species", "the same"),
          tag_size = 7)
```

# exercise: tree representation

```{r}
p1 <- ggtree(portal_tree, layout="ape")
p2 <- ggtree(portal_tree, layout="rectangular")
p3 <- ggtree(portal_tree, layout="roundrect")
p4 <- ggtree(portal_tree, layout="slanted")
p5 <- ggtree(portal_tree, layout="ellipse")
p6 <- ggtree(portal_tree, layout="dendrogram")
p7 <- ggtree(portal_tree, layout="circular")
p8 <- ggtree(portal_tree, layout="radial")
p9 <- ggtree(portal_tree, layout="fan", open.angle = 90)
p10 <- ggtree(portal_tree, layout="equal_angle")
p11 <- ggtree(portal_tree, layout="daylight")
p12 <- ggtree(portal_tree, layout="unrooted")
```


part 2:

create a subplot for each tree

```{r}
plot_list(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, tag_levels = "a")
```


part 3:

is the tree representation rooted or unrooted

```{r}
my_labels <- paste(letters[1:12], "=", "rooted")
my_labels
plot_list(p1, p2, p3, p4, p5, p6, p7, p8, p9, p10, p11, p12, 
          labels = my_labels, tag_size = 5) 
```



```{r}
taxonomy <- read.csv(file = "../raw_data/portal-species-taxonomy.csv")
species <- read.csv(file = "../raw_data/species.csv")
surveys <- read.csv(file = "../raw_data/surveys.csv")
```


join the data table to a tree

```{r}
intersect(colnames(species), colnames(surveys))
library(dplyr)
inner_join(species, surveys, by = "species_id")
```


```{r}
left_join(portal_tree, taxonomy, by = "label")
head(taxonomy)
taxonomy$label
```


replace the tip labels from the original portal tree with a new vector of tip labels
```{r}
tip_labels <- portal_tree$tip.label
tip_labels
tip_labels_cleaned <- gsub("_", " ", tip_labels)
portal_tree[[3]]
portal_tree[["tip.label"]] <- tip_labels_cleaned
portal_tree$tip.label
```


```{r}
tree_data <- left_join(portal_tree, taxonomy, by = "label")
tree_data
class(tree_data)
```

color by a variable in our column:
```{r}
ggtree(tr = tree_data, mapping = aes(col = taxa)) +
  geom_tiplab(size = 2.5, fontface = "italic") +
  xlim(0,100)
ggsave(filename = "../fig/portal-tree-1.png")
```



```{r}
portal_tree <- read.tree(file = "https://raw.githubusercontent.com/LunaSare/data-science-research-biology/main/data/portal-tree.tre")

write.tree(phy = portal_tree, file = "../raw_data/portal-tree-url.tre")

taxonomy <- read.csv(file = "https://raw.githubusercontent.com/LunaSare/data-science-research-biology/main/data/portal-species-taxonomy.csv")

head(taxonomy)
write.csv(taxonomy, file = "../raw_data/taxonomy-test.csv", row.names = FALSE)
```

### join taxonom,y data with portal tree

```{r}
install.packages("tidytree")
library(tidytree)
```

```{r}
library(ggtree)
ggtree(tr = tree_data, mapping = aes(col = genus)) +
  geom_tiplab(size = 2.5, fontface = "italic") +
  xlim(0,100)
ggsave(filename = "../fig/portal-tree-1.png")
```


###  troubleshooting treee and data joins:

```{r}
portal_tree$tip.label
taxonomy$label
match(taxonomy$label, portal_tree$tip.label)
match(portal_tree$tip.label, taxonomy$label)
```


### exercise - homework


```{r}
small_tree <- read.tree(file = "http://ape-package.ird.fr/APER/APER2/primfive.tre")
ggtree(small_tree) +
  geom_tiplab() +
  xlim(0,1.5) +
  geom_tiplab(fontface = "italic")

label <- small_tree$tip.label[1:2]
taxa <- c("Hominini", "Hominidae")
common_name <- c("Human", "Orangutan")

taxon_human_orang <- data.frame(label, taxa, common_name)

```


```{r, eval=FALSE}
install.packages("TDbook")
```

## mapping data ot the tree structure

load data from package TDbook

```{r}
library(TDbook)
data("tree_boots")
data("df_tip_data")
data("df_inode_data")
small_tree$tip.label

tree_boots
names(tree_boots)

#change newick label to just "label"

df_tip_data
names(df_tip_data)

df_inode_data
names(df_inode_data)

```


use mutate
```{r}
library(dplyr)
df_tip_data %>% 
  mutate(label = Newick_label) -> renamed_tip_data
colnames(renamed_tip_data)


df_inode_data %>% 
  mutate(label = newick_label) -> renamed_inode_data
renamed_inode_data$label
```


using colnamed fucntion

```{r}
colnames(df_tip_data)[1] <- "label"
colnames(df_tip_data)

colnames(df_inode_data)[1] <- "label"
colnames(df_inode_data)
```

join tree_boots to df_tip_data:

```{r}
intersect(names(tree_boots), colnames(df_tip_data))
tree_joined <- left_join(tree_boots, df_tip_data, by = "label")
tree_joined
```


plot some data to the tip of tree_boots

```{r}
colnames(df_tip_data)
df_tip_data$rank

ggtree(tree_joined) +
  geom_tippoint(mapping = aes(shape = trophic_habit, col = trophic_habit,
                              size = mass_in_kg)) +
  geom_tiplab(fontface = "italic") +
  xlim(0, 4)

df_tip_data

```

# homwework finish small tree and do mapping weight data exercise

```{r}
library(dplyr)
colnames(surveys)
surveys %>% 
  filter(!is.na(weight)) %>% 
  group_by(species_id) %>% 
  summarize(mean_weight = mean(weight)) -> species_weight

nrow(species_weight)
spp_weight <- inner_join(species, species_weight)
  
colnames(taxonomy)
taxonomy_weight <- full_join(taxonomy, spp_weight)
taxonomy_weight

tree_join <- left_join(portal_tree, taxonomy_weight, 
                       by = "label")

ggtree(tree_join) +
  geom_tiplab(fontface = "italic", size = 2) +
  xlim(0, 20) +
  geom_tippoint(mapping = aes(shape = "circle", 
                              size = mean_weight, col = taxa))
```



