---
title: "shrub_volume.Rmd"
author: "Marcos Padilla-Ruiz"
date: "2022-10-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
library(dplyr)
```


Exercise 1: data wrangling basics
## shrub data

The shrub volume dataset consists of a table made up of five columns labeled site, experiment, length, width, and height with a total of 12 rows


## read data

```{r}
shrubs <- read.csv(file = "../raw_data/shrub-volume-data.csv")
```


## selecting data

```{r}
select(shrubs, length)

select(shrubs, site, experiment)
```


## adding clomun using mutate function

```{r}
mutate(shrubs, area = length*width)
```


## sort data by length with arrange function
```{r}
arrange(shrubs, length)
```



## filter data 

```{r}
filter(shrubs, height > 5)
# shrubs with height greater than 5

filter (shrubs, height > 4, width > 2)
# shrubs with height greater than 4 and width greater than 2

filter(shrubs, experiment == "1" | experiment == "3")
# shrubs from experiment 1 or 3

filter(shrubs, !is.na(height))
# filter to remove null values from height column
```



## new shrub data frame with added column

```{r}
shrub_volumes <- mutate(shrubs, volumes = length*width*height)

shrub_volumes
```


Exercise 2: data aggregation

## rewrite code as pipeline for maximum plant height
```{r}
avg_height <- shrubs |>
  group_by(site) |>
  summarize(avg_ht = max(height)) 
  

avg_height
```


Exercise  3: fix the code

```{r}
shrubs <- read.csv(file = "../raw_data/shrub-volume-data.csv") # they did not specify the location of the file nor did they assign it to a data frame


shrub_data_site <- shrubs |> # 'shrub_data_site' is the name of the new data frame we're creating and 'shrubs' is the data set we are using
  mutate(volume = length * width * height) |> # this line creates a new column labeled 'volume' and is quantified by three columns in the data multiplied with one another
  group_by(site) |> # this function allows us to group the rows by the values in the site column
  summarize(mean_volume = max(volume)) # this summarizes the maximum volumes in each site into one new column

shrub_data_site


shrub_data_exp <- shrubs |> # 'shrub_data_exp' is the new data frame and we are once again working with the 'shrubs' data
  mutate(volume = length * width * height) |> # creates new 'volume' column
  group_by(experiment) |> # groups rows by the values in the experiment column
  summarize(mean_volume = mean(volume)) # summarizes the average volumes within each experiment

shrub_data_exp
```





