---
title: "shrub_volume.Rmd"
author: "Marcos Padilla-Ruiz"
date: "2022-10-18"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
library(dplyr)
```


Exercise 1: data wrangling basics
## shrub data

The shrub volume dataset consists of a table made up of five columns labeled site, experiment, length, width, and height with a total of 12 rows


## read data

```{r}
shrubs <- read.csv(file = "../raw_data/shrub-volume-data.csv")
```


## selecting data

```{r}
select(shrubs, length)

select(shrubs, site, experiment)
```


## adding clomun using mutate function

```{r}
mutate(shrubs, area = length*width)
```


## sort data by length with arrange function
```{r}
arrange(shrubs, length)
```



## filter data 

```{r}
filter(shrubs, height > 5)
# shrubs with height greater than 5

filter (shrubs, height > 4, width > 2)
# shrubs with height greater than 4 and width greater than 2

filter(shrubs, experiment == "1" | experiment == "3")
# shrubs from experiment 1 or 3

filter(shrubs, !is.na(height))
# filter to remove null values from height column
```



## new shrub data frame with added column

```{r}
shrub_volumes <- mutate(shrubs, volumes = length*width*height)

shrub_volumes
```


Exercise 2: data aggregation

## rewrite code as pipeline for maximum plant height
```{r}
avg_height <- shrubs |>
  group_by(site) |>
  summarize(avg_ht = max(height)) 
  

avg_height
```


Exercise  3: fix the code

```{r}
shrubs <- read.csv(file = "../raw_data/shrub-volume-data.csv") # they did not specify the location of the file nor did they assign it to a data frame


shrub_data_site <- shrubs |> # 'shrub_data_site' is the name of the new data frame we're creating and 'shrubs' is the data set we are using
  mutate(volume = length * width * height) |> # this line creates a new column labeled 'volume' and is quantified by three columns in the data multiplied with one another
  group_by(site) |> # this function allows us to group the rows by the values in the site column
  summarize(mean_volume = max(volume)) # this summarizes the maximum volumes in each site into one new column

shrub_data_site


shrub_data_exp <- shrubs |> # 'shrub_data_exp' is the new data frame and we are once again working with the 'shrubs' data
  mutate(volume = length * width * height) |> # creates new 'volume' column
  group_by(experiment) |> # groups rows by the values in the experiment column
  summarize(mean_volume = mean(volume)) # summarizes the average volumes within each experiment

shrub_data_exp
```


Homework Exercises

## Exercise 4

```{r}

shrub_vol_exp <- read.csv(file = "../raw_data/shrub-volume-experiments.csv")

shrub_vol_sites <- read.csv(file = "../raw_data/shrub-volume-sites.csv")


```


## Exercise 5

```{r}
length <- c(2.2, 2.1, 2.7, 3.0, 3.1, 2.5, 1.9, 1.1, 3.5, 2.9)
width <- c(1.3, 2.2, 1.5, 4.5, 3.1, NA, 1.8, 0.5, 2.0, 2.7)
height <- c(9.6, 7.6, 2.2, 1.5, 4.0, 3.0, 4.5, 2.3, 7.5, 3.2)

#1
min(length, na.rm = TRUE)
min(width, na.rm = TRUE)
min(height, na.rm = TRUE)

#2
max(length, na.rm = TRUE)
max(width, na.rm = TRUE)
max(height, na.rm = TRUE)

#3
sum(length, na.rm = TRUE)
sum(width, na.rm = TRUE)
sum(height, na.rm = TRUE)

#4
mean(length, na.rm = TRUE)
mean(width, na.rm = TRUE)
mean(height, na.rm = TRUE)

#5
shrub_volume <- length*width*height

#6
sum(shrub_volume, na.rm = TRUE)

#7
shrub_2.5 <- height[c(length > 2.5)]

#8
shrub_5 <- height[c(height > 5)]

#9 
height[1:5]

#10
shrub_volume[1:3]

#11
tail(shrub_volume)
```

## Exercise 6

```{r}
shrub_dimensions <- read.csv(file = "../raw_data/shrub-dimensions-labeled.csv")

colnames(shrub_dimensions)

#1
shrub_dimensions$length

#2
shrub_dimensions[, c()]

#3
shrubID_and_height <- shrub_dimensions[, c("shrubID", "height")]

#4 
second_row_shrubs <- shrub_dimensions[2,]

#5
first_five_rows_shrubs <- shrub_dimensions[1:5,]

first_five_rows_shrubs
```








