---
title: "small mammals dataset"
author: "Marcos Padilla-Ruiz"
date: "2022-10-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
library(dplyr)
```

```{r}
surveys <- read.csv(file = "../raw_data/surveys.csv")
plots <- read.csv(file = "../raw_data/plots.csv")
species <- read.csv(file = "../raw_data/species.csv")
```

```{r}
select(surveys, year, species_id, weight)
surveys_short <- select(surveys, year, species_id, weight)
head(surveys_short)
```


other ways to subset
```{r}
columns2keep <- c("year", "species_id", "weight")
surveys[,columns2keep]
```


```{r}
colnames(surveys)
surveys$weight_decimal <- surveys$weight/10
```

```{r}
surveys <- mutate(surveys, weight_decimal_2 = weight/10)
```

```{r}
surveys
```


## reorder/rearrange a data table by some value

we will use the function `arange()`

```{r}
arrange(surveys, weight)
```

to arrange in a descending order:
```{r}
arrange(surveys, desc(weight))
```


```{r}
surveys_arranged <- arrange(surveys, weight)

tail(surveys_arranged, 100)
```


## filter data

to filter unwanted data or by some specific value in our data we use the function `filter()`

```{r}
filter(surveys, species_id == "DS")
filter(surveys, species_id != "DS")


```



we an also filter by multiple conditions

```{r}
filter(surveys, species_id == "DS", year > 1991)
filter(surveys, species_id == "DS", year >= 1991)
#above line is "above or equal"
```

the logical operator '&' and '|' mean "and" and "or"

```{r}
temp_table <- filter(surveys, species_id == "DS", year >= 1998)
tail(temp_table, 100)
```

```{r}
filter(surveys, species_id == "DS" & year >= 1998)
```


```{r}
filter(surveys, species_id == "DS" | year >= 1998)

```


lets filter for values that are NA we need the function `is.na()`


```{r}
is.na(surveys$weight)
surveys$weight
filter(surveys, !is.na(weight))

```




1.
```{r}
surveyss <- read.csv(file = "../raw_data/surveys.csv")
```

2.
```{r}
select(surveyss, year, month, day, species_id)

```


3.
```{r}
select(surveyss, "year", "species_id", "weight")
columns <- c("year", "species_id", "weight")
surveyss[,columns]
```


```{r}

surveyss <- filter(surveys, !is.na(weight))
surveyss <- mutate(surveyss, weight_kg = weight/1000)

head(surveyss)
filter(surveyss, species_id == "SH")
```


There are a total of 141 rows after filtering by "SH"



`%>%` - magrittr package required
`|>` - inbuilt in R, very new so you need an updated version

```{r}
mean(x = c(1,3,65, 38, 100))
c(1, 3, 65, 38, 100) %>% mean()

sqrt(mean(x = c(1,3,65,38, 100)))
c(1,3,65,38, 100) |> mean() |> sqrt()

surveys$weight |> mean(na.rm = TRUE)
```



exercise 2

1.
```{r}
surveys1 <- surveys |>
  mutate(weight_kg = weight/1000) |>
  select(year, species_id, weight_kg) |>
  filter(!is.na(weight_kg))
is.na(surveys1$weight_kg) |> any()
```
2.
```{r}

surveys2 <- surveys |>
  filter(species_id == "SH") |>
  filter(!is.na(weight)) |>
  select(year, month, day, species_id)
```

```{r}
ds_weight_by_year <- surveys |>
  filter(species_id == "DS") |>
  filter(!is.na(weight)) |>
  arrange(year) |>
  select(year, weight, species_id)



head(ds_weight_by_year)
nrow(ds_weight_by_year)
colnames(ds_weight_by_year)
is.na(ds_weight_by_year$weight) |> any()
ds_weight_by_year$species_id
```


## data aggregation

```{r}
surveys_by_year <- group_by(surveys, year, plot_id)

surveys_by_year

```



```{r}
summarize(surveys_by_year, abundance = n())
```


the same but using a pipe

```{r}
plot_year_counts <- surveys |>
  group_by(plot_id, year) |>
  summarize(abundance = n())

plot_year_counts

```




exercise 3

1.
```{r}
individuals_species_id <- group_by(surveys, species_id)
summarize(individuals_species_id, abundance = n())
```

2.
```{r}
species_id_and_year <- group_by(surveys, species_id, year)
summarize(species_id_and_year, abundance = n())
```


3.
```{r}
species_DO <- filter(surveys, species_id == "DO", !is.na(weight))
group_species_year_weight <- group_by(species_DO, year)
summarize(group_species_year_weight, average_mass = mean(weight))
```





how to use funciton filter()
some examples of relationl statments: they evaluate reltionships between two varibales or objects
symbols we use are
== "equals to"
> "larger than"
< "smaller than"
!= "not equal to"
>= "larger or equal to"
<= "smaller or equal to"


```{r}
"a" == "a"
"marcos" < "Marcos"
```


logical statements test if two or more relational sttaements are true or false in combination
to evaluate a logical comparison we have two symbols. two types of logical comparisons
- the ampersand &: this evalues thta the relation to the left AND the relation to the right are both TRUE
- the vertical bar |: evaluates that the relation to the left OR the relation to the right are TRUE; in other words at least on these relations must be true

```{r}
1 == 1 & 1 < 2
1 == 1 | 1 < 2
1 == 1 & 1 < 2
```

```{r}
combined <- inner_join(surveys, species, by = "species_id")
combined_full <- inner_join(combined, plots, by = "plot_id")
filter(combined_full, plot_type == "Control" & species_id == "PP" & 
         plot_id == 2 | species_id == "DO" & !is.na(weight)) 
```


```{r}
surveys$weight %>% 
  mean(na.rm = TRUE)

```

```{r}
surveys %>% 
  filter(!is.na(weight)) %>% 
  lm(weight ~ year, data = .)
```





something importnant for relational statements is the NA
```{r}
NA == 3

is.na(NA)
is.na("marcos")

mean(surveys$weight, na.rm = TRUE)
min(surveys$weight, na.rm = TRUE)
max(surveys$weight, na.rm = TRUE)
```




## joining data tables

join data by 

```{r}
inner_join(surveys, species, by = "species_id")
left_join(surveys, species, by = "species_id")
right_join(surveys, species, by = "species_id")
full_join(surveys, species, by = "species_id")
```


how to find out if we have matching columns in two tables
```{r}
surveys_colnames <- colnames(surveys)
species_colnames <- colnames(species)
surveys_colnames[7]
to_keep <- c(1, 4, 3, 2)
to_keep <- c(FALSE, TRUE, TRUE, FALSE)
species_colnames[to_keep]
```


the %in% operator

```{r}
1 %in% c(1, 2, 3)
c(1, 2, 3) %in% 1

matching_colnames <- colnames(species) %in% colnames(surveys)
matching_colnames
colnames(species)[matching_colnames]
```


how to join more than two tables
```{r}
combined <- inner_join(surveys, species, by = "species_id")
combined_full <- inner_join(combined, plots, by = "plot_id")
head(combined)
head(combined_full)
```

how do we run the code using a pipe
```{r}
combined_full <- inner_join(surveys, species, by = "species_id") %>%
  inner_join(plots, by = "plot_id")
head(combined_full)
```


joint-exercise:

```{r}
surveys |>
  inner_join(plots, by = "plot_id") |>
  filter(plot_type == "Control")
  
```

exercise 2:
```{r}
colnames(surveys)
combined_full <- inner_join(surveys, species, by = "species_id") |>
  inner_join(plots, by = "plot_id")
colnames(combined_full)
combined_full %>% 
  select(year, genus, species, weight, plot_type, taxa) |>
  filter(plot_type == "Control" | plot_type == "Long-term Krat Exclosure") %>% 
  filter(taxa == "Rodent") %>% 
  select(year, genus, species, weight, plot_type) %>% 
  filter(!is.na(weight))

```

```{r, eval = FALSE}
#1
combined_full %>% 
  filter(species_id == "DO") %>% 
  select(year, month, day, species_id, weight) -> data1

#2
combined_full %>% 
  filter(species_id == "PP" | species_id == "PB") %>% 
  filter(year >= 1995) %>% 
  filter(!is.na(hindfoot_length)) %>% 
  select(year, species_id, hindfoot_length) -> data2

#3
combined_full %>% 
  group_by(species_id, year) %>% 
  filter(!is.na(hindfoot_length)) %>% 
  summarize(avg_hindfoot_length = mean(hindfoot_length, na.rm = TRUE)) 
  
#4
combined_full %>% 
  select(year, genus, species, weight) %>% 
  filter(genus, plot_type == "Dipodomys")

#5
library(ggplot2)
ggplot(combined_full, mapping = aes (x = weight, y = hindfoot_length,
                                     color = species_id)) +
  labs(x = "Weight", y = "Hindfoot Length", title = "Species by Weight & Hindfoot Length") +
  geom_point(size = 1, alpha = 0.2) +
  scale_x_log10()

#6
ggplot(data = combined_full, mapping = aes(x = weight, fill = species_id)) +
  geom_histogram(bins = 9, col = "black") +
  labs(x = "Weight", y = "Number of Individuals", title = "Species ID Histogram") +
   facet_wrap(~species_id, scales = "free_y")

#7
ggplot() +
  geom_histogram(data = combined_full, 
                 mapping = aes(x = weight, y = "PP", col = species_id)) 
```


```{r}
library(magrittr)
library(dplyr)
surveys %>% 
  filter(species_id == "DM" | species_id == "DO" | species_id == "DS") %>% 
  select(month, day, year, species_id, weight, hindfoot_length) %>%
  filter(!is.na(weight) & !is.na(hindfoot_length)) %>% 
  mutate(species_id = tolower(species_id))


```




```{r, child = 'vectors_and_data_frames.Rmd'}
```


