---
title: "small mammals dataset"
author: "Marcos Padilla-Ruiz"
date: "2022-10-11"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
library(dplyr)
```

```{r}
surveys <- read.csv(file = "../raw_data/surveys.csv")
plots <- read.csv(file = "../raw_data/plots.csv")
species <- read.csv(file = "../raw_data/species.csv")
```

```{r}
select(surveys, year, species_id, weight)
surveys_short <- select(surveys, year, species_id, weight)
head(surveys_short)
```


other ways to subset
```{r}
columns2keep <- c("year", "species_id", "weight")
surveys[,columns2keep]
```


```{r}
colnames(surveys)
surveys$weight_decimal <- surveys$weight/10
```

```{r}
surveys <- mutate(surveys, weight_decimal_2 = weight/10)
```

```{r}
surveys
```


## reorder/rearrange a data table by some value

we will use the function `arange()`

```{r}
arrange(surveys, weight)
```

to arrange in a descending order:
```{r}
arrange(surveys, desc(weight))
```


```{r}
surveys_arranged <- arrange(surveys, weight)

tail(surveys_arranged, 100)
```


## filter data

to filter unwanted data or by some specific value in our data we use the function `filter()`

```{r}
filter(surveys, species_id == "DS")
filter(surveys, species_id != "DS")


```



we an also filter by multiple conditions

```{r}
filter(surveys, species_id == "DS", year > 1991)
filter(surveys, species_id == "DS", year >= 1991)
#above line is "above or equal"
```

the logical operator '&' and '|' mean "and" and "or"

```{r}
temp_table <- filter(surveys, species_id == "DS", year >= 1998)
tail(temp_table, 100)
```

```{r}
filter(surveys, species_id == "DS" & year >= 1998)
```


```{r}
filter(surveys, species_id == "DS" | year >= 1998)

```


lets filter for values that are NA we need the function `is.na()`


```{r}
is.na(surveys$weight)
surveys$weight
filter(surveys, !is.na(weight))

```




1.
```{r}
surveyss <- read.csv(file = "../raw_data/surveys.csv")
```

2.
```{r}
select(surveyss, year, month, day, species_id)

```


3.
```{r}
select(surveyss, "year", "species_id", "weight")
columns <- c("year", "species_id", "weight")
surveyss[,columns]
```


```{r}

surveyss <- filter(surveys, !is.na(weight))
surveyss <- mutate(surveyss, weight_kg = weight/1000)

head(surveyss)
filter(surveyss, species_id == "SH")
```


There are a total of 141 rows after filtering by "SH"



`%>%` - magrittr package required
`|>` - inbuilt in R, very new so you need an updated version

```{r}
mean(x = c(1,3,65, 38, 100))
c(1, 3, 65, 38, 100) %>% mean()

sqrt(mean(x = c(1,3,65,38, 100)))
c(1,3,65,38, 100) |> mean() |> sqrt()

surveys$weight |> mean(na.rm = TRUE)
```



exercise 2

1.
```{r}
surveys1 <- surveys |>
  mutate(weight_kg = weight/1000) |>
  select(year, species_id, weight_kg) |>
  filter(!is.na(weight_kg))
is.na(surveys1$weight_kg) |> any()
```
2.
```{r}

surveys2 <- surveys |>
  filter(species_id == "SH") |>
  filter(!is.na(weight)) |>
  select(year, month, day, species_id)
```

```{r}
ds_weight_by_year <- surveys |>
  filter(species_id == "DS") |>
  filter(!is.na(weight)) |>
  arrange(year) |>
  select(year, weight, species_id)



head(ds_weight_by_year)
nrow(ds_weight_by_year)
colnames(ds_weight_by_year)
is.na(ds_weight_by_year$weight) |> any()
ds_weight_by_year$species_id
```


## data aggregation

```{r}
surveys_by_year <- group_by(surveys, year, plot_id)

surveys_by_year

```



```{r}
summarize(surveys_by_year, abundance = n())
```


the same but using a pipe

```{r}
plot_year_counts <- surveys |>
  group_by(plot_id, year) |>
  summarize(abundance = n())

plot_year_counts

```




exercise 3

1.
```{r}
individuals_species_id <- group_by(surveys, species_id)
summarize(individuals_species_id, abundance = n())
```

2.
```{r}
species_id_and_year <- group_by(surveys, species_id, year)
summarize(species_id_and_year, abundance = n())
```


3.
```{r}
species_DO <- filter(surveys, species_id == "DO", !is.na(weight))
group_species_year_weight <- group_by(species_DO, year)
summarize(group_species_year_weight, average_mass = mean(weight))
```







