---
title: "uhuru-dataset-visualization.Rmd"
author: "Marcos Padilla-Ruiz"
date: "2022-10-04"
output: pdf_document
---


### 1. Describing the data we are using


add a picture of an acacia
### 2. reading the data table into R


```{r}
getwd()
acacia <- read.csv(file = "/Users/marcos/Desktop/BIO 197/Data_Science_Project/raw_data/ACACIA_DREPANOLOBIUM_SURVEY (1).txt", sep = "\t")
```

### 3. explore our data

```{r}
head(acacia) 
tail(acacia)
summary(acacia)
colnames(acacia)
nrow(acacia)
```

make sure that everything that  is a number is actually numeric
use functon summary to do this and check that the type of data corresponds to this
nanother wya is to use th type function

```{r}
typeof(acacia[, "HEIGHT"])

acacia$HEIGHT

```

idenitifed a colummn thta has problmeatic data so wee need to fix it so were gonna read the data table and assign "NA" to "dead" value in the height column


```{r}
acacia <- read.csv(file = "/Users/marcos/Desktop/BIO 197/Data_Science_Project/raw_data/ACACIA_DREPANOLOBIUM_SURVEY (1).txt", 
                   sep = "\t", na.strings = "dead")
getwd()
```




```{r}
acacia$HEIGHT
```

### 4. visualize our data

for this we use ggplot package
let install it first


```{r}
#install.packages("ggplot2")
library(ggplot2)
```


now we are gonna creat oiuur first  plotting layer with ggplot function

```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y = HEIGHT)) + 
  geom_point(size = 1, col= "magenta", alpha = 0.2) +
  labs(x = "Circumference", y = "Height", title = "Data from Acacia Survey")
```

to rescale the plotting of the axis to log scale we use the function scale_y_log_10()

```{r}
ggplot(data = acacia, mapping = aes(x = CIRC, y =  HEIGHT)) +
  geom_point(size = 1, col= "cyan4", alpha = 0.2) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Circumference", y = "Height", title = "Data from Acacia Survey")
```

we have the informaiton pon experimental treatmenet in treatmnt column

```{r}
acacia$TREATMENT
```

lets add this information to our plot:

```{r}
ggplot(acacia, mapping = aes (x = CIRC, y = HEIGHT, color = TREATMENT)) +
  geom_point(size = 2, alpha = 0.5) +
  labs(x = "Circumference", y = "Height", title = "Data from Acacia Survey") +
  scale_x_log10() +
  scale_y_log10() 
```

### 4.2 visualize a statistical anlaysis of correlation 







